Exploring probability distributions for bivariate temporal granularities

### 2019-12-02 gravitas/gravitas.Rmd
aperiodic completed

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2[@Wickham2009-pk]. 
bibliography: [bibliography.bib]
output:
  bookdown::pdf_book:
    #base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

@@

```{r external, include = FALSE}
# read_chunk('scripts/main.R')
```

```{r load}

@@

# Linear Time granularities {#sec:lineartime}

<!-- general calendar categorization -->
<!-- Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Classes of time abstractions may be based on arrangement or hierarchical order of time, discussed in further details in \autoref{sec:arrangement} and \autoref{sec:order}. -->

Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Time granularities are **linear** if they respect the linear progression of time. Examples include hours, days, weeks and months.

## Definitions and Relationships

<!-- Arrangement: Linear vs. Circular vs. aperiodic {#sec:arrangement} -->

<!-- The arrangement of the time domain can result in deconstructing time in linear, circular or aperiodic (nearly circular) ways.  -->

@@

<!-- In the following sections, we provide a formalism to all of these abstractions and identify their relationships and symbolic representations.  -->

<!-- ### Linear {#sec:linear-gran-def} -->

@@

  (1) if $i < j$ and $G(i)$ and $G(j)$ are non-empty, then each element of $G(i)$ is less
than all elements of $G(j)$, and  
  (2) if $i < k < j$ and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty.  
\end{timedom}

\begin{timedom}\label{def:granule}
Each non-empty subset $G(i)$ is called a granule, where $i$ is one of the indexes and $G$ is a linear granularity.
\end{timedom}

**Dicussion:** The first condition in \ref{def:linear} implies that the granules in a linear granularity are non-overlapping and their index order is same as time order.
\autoref{fig:linear-time} shows the implication of this condition. It shows the linear granularities hours, days, weeks, months and years, each of which are unidirectional in nature and arranged from past to future. If we consider the chronon [@aigner2011visualization] as hourly, the time domain with T hours will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on. Each granule of the linear granularities shown are represented through a box.

```{r linear-time,echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear granularities"}
knitr::include_graphics("Figs/linear-arr.png")
```

@@

\begin{timedom}\label{def:finerthan}
A linear granularity G is finer than a  linear granularity H, denoted $G \preceq H$, if
for each index i, there exists an index j such that $G(i) \subset H(j)$.
\end{timedom}.

\begin{timedom}\label{def:groupsinto}
A linear granularity G groups into a linear granularity H, denoted
$G \trianglelefteq H,$
if for each index j there exists a (possibly infinite) subset S of the integers such
that
\begin{equation}
H(j) = \bigcup_{i \in S}G(i)
\end{equation}
\end{timedom}

**Dicussion:** $G \trianglelefteq H$ if each granule H(j) is the union of some granules of G.  $G \preceq H$ if every granule in G is a subset of some granule in H. In \autoref{fig:groups-finer}, both $Days \trianglelefteq Weeks$ and $Days \preceq Weeks$ hold true. However, Weekends groups into Holidays but Weekends are not finer than Holidays. On the other hand, Weekends are finer than weeks but do not group into weeks. Consider another example where $Days \trianglelefteq Months$. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions, .

```{r groups-finer,echo=FALSE,out.width = "100%",fig.cap="Illustrations of groups into and finer than relationships. Weekends groups into Holidays but Weekends are not finer than Holidays. On the other hand, Weekends are finer than weeks but do not group into weeks. Day groups into and are finer than weeks"}
knitr::include_graphics("Figs/groupsandfiner.png")
```

\begin{timedom}\label{def:periodical}
A granularity H is periodical with respect to a granularity G if
(1) $G \trianglelefteq H,$, and
(2) there exist $R,P \epsilon Z+$, where R is less than the number of granules of H, such
that for all $i \epsilon Z$, if $H(i) = \bigcup_{j \in S}G(j)$ and $H (i + R) \neq \phi$ then
$H (i + R) = \bigcup_{j \in S} G(j + P)$.
\end{timedom}

A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, ${S_0,...,S_(R-1)}$ describing the granules of H within one period;
(ii) the value of P; 
(iii) the indexes of first and last granules in H, if their value is not infinite. 
Then, if ${S_0,...,S_{R-1}}$ are the sets of indexes of G describing
$H (0), . . . , H (R - 1)$, respectively, then the description of an arbitrary granule H(j) is given by: $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$.

**Dicussion:**

@@

\begin{timedom}\label{def:quasi}
A granularity H is quasi-periodical with respect to a granularity G if
(1) $$G \trianglelefteq H,$$, and
(2) there exist a set of intervals E1,...,Ez (the granularity exceptions) and positive
integers R,P, where R is less than the minimum of the number of granules of H
between any 2 exceptions, such that for all $i \epsilon Z$, if $H(i) = \bigcup_{k \in [0,k]} G(j_r)$  and $H (i + R) \neq \phi$ and $i + R < min(E)$, where E is the closest existing exception after $H (i)^2$, then $H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P )$.
\end{timedom}

**Dicussion:**

@@

\begin{timedom}\label{def:BG}
Bottom granularity - Given a granularity order relationship g-rel and a set of granularities having the same time domain, a granularity G in the set is a bottom granularity with respect to g-rel, if G g-rel H for each granularity H in the set. 
\end{timedom}

**Dicussion:** Given the set of all granularities defined over the time domain (Z; <), and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

@@

## Computation through Calendar Algebra

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra [@Ning2002-tf]. It is assumed that there exists a “bottom” granularity and Calendar algebra operations are designed to generate
new granularities from the bottom one or recursively, from those already generated.
Thus, the relationship between the operand(s) and the resulting granularities are encoded
in the operations.

@@

- **The grouping operation** : Let G be a full-integer labeled granularity, and m a positive integer. The grouping operation $Group_m(G)$ generates a new granularity G, by partitioning the granules of G into m-granule groups and making each group a granule of the resulting granularity.
More precisely, $G = Group_m(G)$ is the full-integer labeled granularity such that for
each integer $i$, $G(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G(j)$.

**Examples** :

@@

- **The altering-tick operation** : Let G1, G2 be full-integer labeled granularities, and $l$, $k$, $m$ integers, where G2 partitions G1, and $1 \leq l \leq m$. The altering-tick operation $Alter^{m}_{l,k} (G2, G1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the $l$th granule of each group has $|k|$ additional (or fewer when k < 0) granules of G2.

@@

For more variations of grouping operations and granule oriented operations, the readers are remcommended to see [@Ning2002-tf].s

# Cyclic Time granularities {#sec:circular-gran-def}

<!-- sub: regular and irregular mapping leads to circular and quasi-circular granularities -->

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. Lining up with the the assumption in [@Bettini2000-vy] that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists and is represented by the index of the tsibble. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceeded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

Cyclic time granularities which accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month. The formalisms would differ for circular and quasi-circular granularities although both accomodate periodicities in data.

<!--sub:How to relate linear granularities to tsibble and manipulation of indices required for circular and quasi-circular granularities -->

## Circular {#sec:circular-gran-def}

@@

\begin{equation} \label{eq:eq2}
C_{BG, G}(z) = BG(z\mod k(BG,G)) \forall z \in \mathbb{Z}^+,
BG(z) \neq \phi,
\end{equation}
where,  
z denotes the index set, 
BG denotes the index of the tsibble (bottom granularity), 
G is a linear granularity periodic with respect to BG with regular mapping,     
and k(BG,G) be the period length of the bottom granularity with respect to linear granularity G.
\end{timedom}

@@

Table \ref{tab:definitions} shows representation of circular granularities relating two linear granularities. It is possible that none of these two linear granularities are bottom granularities. But the representation of the resultant circular granularity will be a function of the bottom granularity. Suppose the bottom granularity is minutes, and $k_i$ is the number of granules of the finer granularity in each granule of the coarser granularity in the circular granularity $C_i$, also referred to as the period length.

@@

## Quasi-circular {#sec:circular-gran-def}

A **quasi-circular** granularity can not be defined using modular arithmetic due to its irregular mapping with the bottom granularity. For example, \autoref{tab:quasi} shows some example of quasi-circular granularities ($QC_i$) and plausible choices of number of granules ($k_i$) of the finer granularity inside each granule of the coarser one in the cyclic granularity ($QC_i$).

@@

Day-of-Month: & $QC_1(z)$ && $k_1=31, 30, 29, 28$\\
Hour-of-Month: & $QC_2(z)$ && $k_2=24*31, 24*30, 24*29, 24*28$\\
Day-of-Year: & $QC_3(z)$ && $k_3=366, 365$\\
Week-of-Month: & $QC_4(z)$ && $k_4=5, 4$\\ 
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative quasi-circular granularities with potential period lengths}
\label{tab:quasi}
\end{table}

\begin{timedom}\label{def:quasicircular}
A quasi-circular granularity QC_(BG, G') that relates linear granularities G' and bottom granularity BG, if

\begin{equation}\label{eq:eq3}
QC_{BG, G'}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert), z \in T_k
\end{equation}

where,  
z denotes the index set,  
BG denotes the index of the tsibble (bottom granularity),  
G' is a linear granularity periodic with respect to BG with irregular mapping,   
R' denotes the number of granules of G' in each period of the grouping (BG,G),   
T_w are the sets of indices of BG describing G'(w) such that $G'(w) = \bigcup_{z \in T_w}BG(z)$,$T_w = 0$ for $w \epsilon Z_{<0}$ and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{timedom}

Since G' is periodical with respect to BG, from definition \ref{def:periodical} the following holds true:

- $BG \trianglelefteq G'$
- $G'(w) = \bigcup_{z \in T_w \mod R'}BG(P'*\lfloor w/R' \rfloor + z)$,  where P' denotes the period of the grouping $(BG, G')$.

**Discussion**: Example showing quasi-circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:quasi-gran}. Each granule is represented by a box. Similar to \autoref{fig:circular-gran}, the granules that overlap share elements from the underlying time domain and the first slot shows the index set and the bottom granularity. 
Two linear granularities AG and AH are considered, where $G' = Alter^{2}_{(1,-1)}(BG, Group_3(BG))$ and $H' = Alter^{2}_{(1,-2)}(BG, Group_7(BG))$. This implies that G' is made up by shrinking every 1st granule of Group_3(BG) by 1 granule and H' is made up of shrinking every 1st granule of Group_3(BG) by 2 granules. Number of granules of G' and H' in each period of BG is 2, but the number of granules of BG in each of those granules are different. QC(BG, G') and QC(BG, H') are repetitive categorization of time, similar to circular granularities, except that the number of granules of BG is not necessarily the same across different granules of G' or H'. For G', R' = 2, P'= 5, $T_0$ = {0,1} and $T_1$ = {2, 3, 4}. For H', R' = 2, P'= 12, $T_0$ = {0, 1, 2, 3, 4, 5, 6} and $T_1$ = {7, 8, 9, 10, 11}, then we will have \autoref{eq:eq4} and  \autoref{eq:eq5}.

\begin{equation} \label{eq:eq4}
\begin{split}
QC(BG,G')(8) & = BG(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) , 8 \in T_{3} \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = BG(8 - 2*\vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert) \\
  & = BG(8 - 2*2 - 3) \\
  & = BG(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq:eq5}
\begin{split}
QC(BG,H')(10) & = BG(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,  10 \in T_{1}  \\
  & = BG(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = BG(10 - \vert T_{0 \mod 2}\vert) \\
  & = BG(10 - \vert T_{0}\vert) \\
  & = BG(10 - 7) \\
  & = BG(3) \\
\end{split}
\end{equation}

```{r quasi-gran, echo=FALSE,out.width = "100%",fig.cap="Relating circular granularities and bottom linear granularity "}
knitr::include_graphics("Figs/aperiodic-granv1.png")
```

@@

## Aperiodic {#sec:circular-gran-def}

Periodic and Quasi-periodic behavior can be defined by a repeating pattern, except for a finite number of granules that can be seen as discontinuity points in the granularity in case of quasi-periodic behavior. Aperiodic time granularities are the ones which can not be specified as a periodic repetition of a pattern of granules. Most public holidays repeat every year, but there is no period within which their behavior remains constant. For example, U.S labour day is the first Monday in September and U.S. Memorial Day is the last day in May. A classic example can be that of Easter, whose dates repeat only after 5,700,000 years. Examples of aperiodic granularity may also include school holidays or a scheduling event that covers the first and third Monday of the month between June and October, except for state holidays. All of these are still recurring events, but recurring with non-periodic patterns. As such, plausible $k_i$ from \autoref{tab:aperiodic} can be infinite for aperiodic granularities.

\begin{timedom}\label{def:acyclic}
An aperiodic granularity A_(BG, G'') that relates linear granularities G and bottom granularity BG, with G'' being aperiodical with respect to BG, is given by

\begin{equation}\label{eq:eq}
\begin{split}
A_{BG, G}(z) & = 1, z \in T_q\\
     & = 0, z \not\in T_q\\
\end{split}
\end{equation}
where,  
z denotes the index set (index of BG),
$T_q$ are the sets of indices of BG describing $G''(q)$ such that $G''(q) = \bigcup_{z \in T_q}BG(z)$,$T_q = 0$ for $q \epsilon Z_{<0}$.
\end{timedom}

@@

# Computing cyclic time granularities {#sec:order}

@@

# Data structure

<!-- How volumes of continuously collected data can be visualized to ease further decision making?  -->

Effective exploration or good visualization require good data structures. Temporal data problems are typically grouped into two types of analysis, time series and longitudinal. A new tidy data structure to support exploration and modeling of temporal data is tsibble, where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. This opens the door to easily operating on time to make calculations and extract different temporal components, such as month and day of the week. Commonly, simple sequences of time value pairs $<t_0, v_0> \dots, <t_n, v_n>$ are the basis of analysis and visualization. It is recognized that the initial approaches of just considering time as an ordinal dimension in a 2D or 3D visualisation are inadequate to capture characteristics of time-dependent information. For example, there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use different calendars that define meaningful system of granularities for different application domain. Data can then be integrated at different levels of granularity enabling value aggregation or distribution along granularities. Since, we are essentially concerned with detection of previously unknown periodic behavior of data, it makes sense to support the detection of patterns by anabling aggregation or distribution across cyclic time granularities. We consider the following data structure for exploring distribution of a measured variable ((Figure \@ref(fig:data-structure)). These are based on conditioning the values taken on by one or more of the cyclic granularities in a data set. This means data subsets are formed for each level of the conditioned variable and the fixed variable. This can be an useful data structure for exploratory data analysis as we can rapidly compare patterns in different parts of the data.

```{r data-structure, fig.cap = "The data structure for exploring periodicities along cyclic granularities. Two cyclic granularities serve as fixed and conditioned variable and the observational unit across combinations of these cyclic time granularities.", fig.pos="htb"}
include_graphics("Figs/data-structure.png")
```

## Fixed and Conditioned variables {#sec:synergy}

The cyclic time granularities which are essentially a function of the index of the tsibble are abstractions of time and hence are discrete variable, mapped as a factor. The relationships between the fixed and conditioned variable are of importance in this data structure due to the ways the cyclic granularities, which are mapped to fixed and conditional variable, relate to one another. To define a general framework, we define harmony and clashes.

@@

<!-- Before exploring the behavior of a "dependent variable" across time granularities, -->
<!-- it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable.  -->

@@

@@

Therefore, pair of cyclic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**. And the ones that do not lead to any missing combinations are called **harmonies**.

## Measured variable

The measured variable which is the univariate observational units can be dicrete or continuous.

<!-- would invlove a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->

 <!-- Instead, the big problem can be broken down into smaller pieces by focusing on specific tasks per visual representation in a 2D space. -->

>>>>

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2[@Wickham2009-pk].

bibliography: bibliography.bib
output: rticles::asa_article
---

@@

@@

# Linear Time granularities{#sec:timegrandef}

<!-- general calendar categorization -->
Often we partition time into months, weeks or days to relate it to data. Such discrete abstractions of time can be thought of as time granularities [@aigner2011visualization]. Examples of time abstractions may also include day-of-week, time-of-day, week-of-year, day-of-month, month-of-year, working day/non-working day, etc which are useful to represent different periodicities in the data. Classes of time abstractions may be based on arrangement or hierarchical order of time, discussed in further details in \autoref{sec:arrangement} and \autoref{sec:order}.

## Arrangement: Linear vs. Circular vs. aperiodic {#sec:arrangement}

The arrangement of the time domain can result in deconstructing time in linear, circular or aperiodic (nearly circular) ways. Time granularities are **linear** if they respect the linear progression of time. Examples include hours, days, weeks and months. However, periodicity is very common in all kinds of data. Time granularities which can accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. Hence, time granularities can also be **circular** or **aperiodic** depending on if the mapping of the linear granularities is regular or irregular. Examples of circular time granularities include hour of the day, and day of the week, whereas examples for aperiodic time granularities can be day of the month or public holidays.

@@

In the following sections, we provide a formalism to all of these abstractions and identify their relationships and symbolic representations.

### Linear {#sec:linear-gran-def}

@@

  (C1) if $i < j$ and $G(i)$ and $G(j)$ are non-empty, then each element of $G(i)$ is less
than all elements of $G(j)$, and  
  (C2) if $i < k < j$ and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty.  
\end{timedom}

**Dicussion:** Each non-empty subset $G(i)$ is called a **granule**, where $i$ is one of the indexes and $G$ is a linear granularity. The first condition implies that the granules in a linear granularity are non-overlapping and their index order is same as time order.
\autoref{fig:linear-time} shows the implication of this condition. If we consider the chronon [@aigner2011visualization] as hourly, the time domain with T hours will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on.

```{r linear-time,echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear granularities"}
knitr::include_graphics("Figs/linear-time.png")
```

#### Relationships between two linear granularities and association to Periodicity

@@

\begin{timedom}\label{def:finerthan}
A linear granularity G is finer than a  linear granularity H, denoted $G \preceq H$, if
for each index i, there exists an index j such that $G(i) \subset H(j)$.
\end{timedom}

\begin{timedom}\label{def:groupsinto}
A linear granularity G groups into a linear granularity H, denoted
G \trianglelefteq H 
if for each index j there exists a (possibly infinite) subset S of the integers such
that
\begin{equation}
H(j) = \bigcup_{i \in S}G(i)
\end{equation}
\end{timedom}

According to this definition, $G \trianglelefteq H$ if each granule H (j) is the union of some
granules of G. For example, $Days \trianglelefteq Months$. However, this  relationship is not fully described until we associate it with periodicity. For example, $Days \trianglelefteq Months,$, since each month is a grouping of a number of days, however it is also a periodical grouping. If leap year is ignored, the periodicity of the grouping is 1 year, since each month would be defined as the grouping of the same number of days (31, 28, or 30, depending on the month) every year. Considering leap years and all their exceptions, the period becomes 400 years. 
\begin{timedom}\label{def:periodical}
A granularity H is periodical with respect to a granularity G if
(1) G \trianglelefteq H, and
(2) there exist R,P \epsilon Z+, where R is less than the number of granules of H, such
that for all i \epsilon Z, if H(i) = \bigcup_{j \in S}G(j) and H (i + R) \neq \phi then
H (i + R) = \bigcup_{j \in S} G(j + P).
\end{timedom}

A granularity H which is periodical with respect to G is specified by:
(i) the R sets of indexes of G, ${S_0,...,S_(R-1)}$ describing the granules of H within one period;
(ii) the value of P; 
(iii) the indexes of first and last granules in H, if their value is not infinite. 
Then, if ${S_0,...,S_{R-1}}$ are the sets of indexes of G describing
$H (0), . . . , H (R - 1)$, respectively, then the description of an arbitrary granule H (j) is given by: $\bigcup_{i \in S_j \mod R}G(P*\lfloor j/R \rfloor + i)$.

This formula can be explained observing that H (j ) is either one of H (0), . . . , H (R - 1) or, for the periodicity of H, is one of these granules “shifted” ahead or behind on the time line of a finite number of periods. If $H (j ) \epsilon {H (0), . . . , H (R - 1)}$, then the formula ‘j mod R’ defines the index (among those in {0,...,R - 1}) of the granule that must be shifted to obtain H (j).The number of periods each granule of G composing H (j mod R) should be shifted is given by $\lfloor j/R \rfloor$.

@@

\begin{timedom}\label{def:quasi}
A granularity H is quasi-periodical with respect to a granularity G if
(1) G \trianglelefteq H, and
(2) there exist a set of intervals E1,...,Ez (the granularity exceptions) and positive
integers R,P, where R is less than the minimum of the number of granules of H
between any 2 exceptions, such that for all i \epsilon Z, if H(i) = \bigcup_{k \in [0,k]} G(j_r)  and H (i + R) \neq \phi and i + R < min(E), where E is the closest existing exception after H (i)^2, then H (i + R) = \bigcup_{k \in [0,k]} G(j_r + P ).
\end{timedom}

@@

\begin{timedom}\label{def:BG}
**Definition: Bottom granularity** Given a granularity order relationship g-rel and a set of granularities having the same time domain, a granularity G in the set is a bottom granularity with respect to g-rel, if G g-rel H for each granularity H in the set. 
\end{timedom}

**Example:** Given the set of all granularities defined over the time domain (Z; <), and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

@@

#### Computation of linear time granularities

Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra[@Ning2002-tf]. It is assumed that there exists a “bottom” granularity and Calendar algebra operations are designed to generate
new granularities from the bottom one or recursively, from those already generated.
Thus, the relationship between the operand(s) and the resulting granularities are encoded
in the operations.

@@

- **The grouping operation** : Let G be a full-integer labeled granularity, and m a positive integer. The grouping operation $Group_m(G)$ generates a new granularity G, by partitioning the granules of G into m-granule groups and making each group a granule of the resulting granularity.
More precisely, $G = Group_m(G)$ is the full-integer labeled granularity such that for
each integer i, $G(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G(j)$.

@@

- **The altering-tick operation** : Let G1, G2 be full-integer labeled granularities, and l, k, m integers, where G2 partitions G1, and $1 \leq l \leq m$. The altering-tick operation $Alter^{m}_{l,k} (G2, G1)$ generates a new full-integer labeled granularity by periodically expanding or shrinking granules of G1 in terms of granules of G2. The altering-tick operation modifies the granules of G1 so that the lth granule of each group has |k| additional (or fewer when k < 0) granules of G2.

@@

### Cyclic {#sec:cyclic-gran-def}

In a cyclic organization of time, the domain is composed of a set of recurring time values. Hence, any time value can be proceeded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A time domain, as defined by Bettini, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. Lining up with the the assumption in [@Bettini2000-vy] that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists and is represented by the index of the tsibble.

The mappings of two linear granularities forming a cyclic granularity can be regular or irregular. It will be regular if, for example, they are formed by grouping operation. Grouping operation would ensure that each granule of the resulting granularity consists of same number of granules of bottom granularity. However, if they are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of bottom granularity. The relationship can still be periodic, depending on the choice of periods and hence these are also useful in exploring periodicities. However, the formalism would differ for circular (regular mapping) and aperiodic (irregular mapping) granularities.

#### Circular {#sec:circular-gran-def}

@@

\begin{equation} \label{eq:eq2}
C_{BG, G}(z) = BG(z\mod k(BG,G)) \forall z \in \mathbb{Z}^+,
BG(z) \neq \phi
\end{equation}
where,  
z denotes the index set.  
BG denotes the index of the tsibble (bottom granularity).  
G be a linear granularity, periodic with respect to BG.    
k(BG,G) be the period length of the bottom granularity with respect to linear granularity G.  
\end{timedom}

@@

Table \ref{tab:definitions} shows representation of circular granularities relating two linear granularities. It is possible that none of these two linear granularities are bottom granularities. But the representation of the resultant circular granularity will be a function of the bottom granularity. For example, suppose the bottom granularity is minutes, and let $k_i$ is the period length of $C_i$.

@@

### Aperiodic {#sec:aperiodic-gran-def}

An **Aperiodic granularity** can not be defined using modular arithmetic. The modulus for these type of calendar categorizations are not constant due to irregular mapping with bottom granularities. For example,\autoref{tab:aperiodic} shows some example of aperiodic granularities ($A_i$) and plausible choices of period lengths($k_i$). The coarse granularity is assumed as the period within which the finer granularity will repeat its behavior aperiodically.

@@

Day-of-Month: & $A_1(z)$ &  $k_1=31, 30, 29, 28$\\
Hour-of-Month: & $A_2(z)$ & $k_2=24*31, 24*30, 24*29, 24*28$\\
Day-of-Year: & $A_3(z)$ & $k_3=366, 365$\\
Week-of-Month: & $A_4(z)$ & $k_4=5, 4$\\ 
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative aperiodic circular granularities with potential period lengths}
\label{tab:aperiodic}
\end{table}

**Definition: Aperiodic granularity** An aperiodic granularity A(AG, BG) that relates linear granularities AG and bottom granularity BG, with AG being periodical with respect to BG, is given by

\begin{equation}\label{eq:eq3}
A_{BG, AG}(z) = BG(z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert), z \in T_k
\end{equation}
where,  
z denotes the index set (index of BG),
R' denotes the number of granules of AG in each period of BG,  
$T_w$ are the sets of indices of BG describing $AG(w)$ such that $AG(w) = \bigcup_{z \in T_w}BG(z)$, and $T_w = 0$ for $w \epsilon Z_{<0}$  
$\vert T_w \vert$ is the cardinality of set $T_w$.

Since AG is periodical with respect to BG, from definition of "Groups periodically into" the following holds true:

- $BG \trianglelefteq AG$
- $AG(w) = \bigcup_{z \in T_w \mod R}BG(P'*\lfloor w/R' \rfloor + z)$,  where P' denotes the period of the granularity BG.

**Discussion**: Example showing aperiodic granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:aperiodic-gran}. Each granule is represented by a box. Similar to \autoref{fig:circular-gran}, the granules that overlap share elements from the underlying time domain and the first slot shows the index set and the bottom granularity. 
Two linear granularities AG and AH are considered, where $AG = Alter^{2}_{(1,-1)}(BG, Group_3(BG))$ and $AH = Alter^{2}_{(1,-2)}(BG, Group_7(BG))$. This implies that AG is made up by shrinking every 1st granule of Group_3(BG) by 1 granule and AH is made up of shrinking every 1st granule of Group_3(BG) by 2 granules. Number of granules of AG and AH in each period of BG is 2, but the number of granules of BG in each of those granules are different. A(BG, AG) and A(BG, AH) are repetitive categorization of time, similar to circular granularities, except that the number of granules of BG is not necessarily the same across different granules of AG or AH. For AG, R' = 2, P'= 5, $T_0$ = {0,1} and $T_1$ = {2, 3, 4}. For AH, R' = 2, P'= 12, $T_0$ = {0, 1, 2, 3, 4, 5, 6} and $T_1$ = {7, 8, 9, 10, 11}, then we will have \autoref{eq:eq4} and  \autoref{eq:eq5}.

\begin{equation} \label{eq:eq4}
\begin{split}
A(BG,AG)(8) & = BG(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) , 8 \in T_{3} \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = BG(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = BG(8 - 2*\vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert) \\
  & = BG(8 - 2*2 - 3) \\
  & = BG(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq:eq5}
\begin{split}
A(BG,AH)(10) & = BG(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,  10 \in T_{1}  \\
  & = BG(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = BG(10 - \vert T_{0 \mod 2}\vert) \\
  & = BG(10 - \vert T_{0}\vert) \\
  & = BG(10 - 7) \\
  & = BG(3) \\
\end{split}
\end{equation}

```{r aperiodic-gran, echo=FALSE,out.width = "100%",fig.cap="Relating circular granularities and bottom linear granularity "}
knitr::include_graphics("Figs/aperiodic-granv1.png")
```

@@

@@

## Order: Single vs. Multiple {#sec:order}

@@

# Synergy of cyclic time granularities {#sec:synergy}

@@

Before exploring the behavior of a "dependent variable" across time granularities,
it is important to know how these granularities interact with each other. If two time granulaties of interest interact, the relationship between each of the interacting variables and the dependent variable will depend on the value of the other interacting variable. To define a general framework, we define harmony and clashes.

@@

## Harmony and Clashes

@@

Therefore, pair of circular/aperiodic granularities which lead to structurally, event-driven or build-based empty-combinations are referred to as to as **clashes**.
And the ones that do not lead to any missing combinations are called **harmonies**.
